Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-08-23 Paper/Paper.tex
images and tables placed at the top of the pages to improve readability of the text
final version, changed *_5.pdf's and some text

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{
 	Fraunhofer IGD, Fraunhoferstr. 5, 64283 Darmstadt, Germany\\
 	\email{tobias.grosse-puppendahl@igd.fraunhofer.de}
 	\and
 	Technische Universit\"at Darmstadt, Hochschulstr. 10, 64289 Darmstadt, Germany\\
 	\email{\{berlin,borazio\}@ess.tu-darmstadt.de}
}

@@

\maketitle
\begin{abstract}
Activity recognition with an accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the user's environment to boost activity recognition. This work presents a sensor prototype consisting of an accelerometer and a capacitive proximity sensor that senses the user's activities based on the combined sensor values. We show that our proposed approach of combining both modalities significantly improves the recognition rate for detecting activities of daily living.
\keywords{activity recognition, capacitive proximity sensors, ambient assisted living, user context}
\end{abstract}

@@

Sensing a person's activity is an active research topic with a raising interest due to the advancement in mobile phone technology. These devices include multiple sensors and therefore enable the recognition of daily activities \cite{brezmes2009activity}. Current wearable activity recognition systems are able to unobtrusively capture and recognize a person's activities throughout the whole day. These systems often rely on inertial sensor data that is captured by wearable sensors embedded in a mobile device \cite{brezmes2009activity} or attached to the body \cite{Ravi2005}. Usually, single sensor modalities are used or duplicated to detect the activities. However, it is a great challenge to identify fine-grained activities just by using a single modality like the accelerometer.

Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the object's distance, its size and the material it is made of. In this work, we show that an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living. Therefore, we obtained an open-hardware and open-source wrist-worn activity data logger \cite{hedgehog} and integrated a capacitive proximity sensor invisibly into the wristband.

There are several intuitive examples for which a combination of accelerometer-based activity recognition with a capacitive proximity sensor reveals its strength. For example, it may be conducted which material is placed underneath the wristband. The capacitive proximity sensor would return a different measurement result for a hand placed on a couch covered with fabric than a hand placed a wooden table. Moreover, the approximate distance from the wristband to objects can be exploited to identify activities like grasping into a locker or a refrigerator to prepare food.

@@

A great variety of capacitive sensors and measurement techniques exists \cite{Smith1999}. The most common sensing principle, the loading mode, is based on running numerous charge and discharge cycles of the virtual capacitor that is created by the electrode and the environment. Depending on the charge and discharge times, one can infer the corresponding capacitance. This sensing principle is applied by Wimmer et al. in \cite{Wimmer2007} who presented a toolkit for capacitive proximity sensing. In previous works, capacitive proximity sensors were applied in various fields of human-computer-interaction. Wimmer et al. and Grosse-Puppendahl et al. presented gesture recognition systems \cite{Wimmer,Grosse-puppendahl2012} as well as smart furniture that can sense human activities \cite{Wimmer} and classify human postures \cite{Grosse-puppendahl2011}. Cheng et al. have investigated the possibility of using capacitive sensors for activity recognition by measuring shape changes of muscles and skin \cite{Cheng2010}. To our knowledge, capacitive proximity sensors have not been embedded into a wearable device to enhance the performance of activity recognition.

@@

The sensor node itself is built around the low-power Microchip microcontroller (PIC18F46J50) featuring an accelerometer sensor (ADXL345) to capture human motion, light and ambient temperature sensors and a microSD flash card for locally storing the sensor data. The sensor is powered by a $200mAh$ lithium polymer battery, which allows for two weeks of continuous recording on a single battery charge. A USB port is used to configure the sensor (e.g. setting the sensitivity of the accelerometer), to access the stored sensor data, and to recharge the battery. A plastic case nicely packages and protects the sensor to be worn at the wrist (Figure \ref{fig:sensornode}).

@@

The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode. The sensor draws a supply current of 1mA at 3.3V when active and qualifies it for wearable proximity sensing applications. In the following a virtual capacitor denotes the capacitance between the sensing electrode and the environment. The capacitance of the sensing electrode to environmental objects increases with closer distances.
% The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode that aims to eliminate the influence of the underlying skin. Without shielding no meaningful measurements could be performed, since the arm's skin is a very proximate grounded object to the sensing electrode. The sensor draws a supply current of 1mA at 3.3V when active and qualifies it for wearable proximity sensing applications. In the following a virtual capacitor denotes the capacitance between the sensing electrode and the environment. The capacitance of the sensing electrode to environmental objects increases with closer distances.

@@

The operations required for a measurement cycle are illustrated in Figure \ref{fig:pseudocode}. The proximity sensor board generates a clock signal with varying frequency depending on the charge and discharge cycles. The HedgeHog measures the resulting capacitance by counting the signal's edges over a gate time of approximately 9.5ms. During that counting phase, the microcontroller is sent to sleep in order to reduce power consumption. In the following, the HedgeHog applies run-length encoding on the measured data to reduce overhead and periodically logs the data to the integrated microSD card.

@@

The recorded test set contains the following activities: opening door, sitting on a couch, lying on a couch, putting kitchen equipments from a shelf and out of a locker, preparing a bread with marmalade, eating the bread, pouring and drinking water, walking and sleeping. The relations of these activities to environmental objects are given in Table \ref{tab:activities}. Some of those activities are very hard to recognize when the data is limited to a single modality like a 3D accelerometer. For example, sitting at the table and sitting on a couch are very similar activities. We aim to show that the data basis can be significantly improved by the additional input modality.

\begin{table}
	\centering
	\caption{Some details on the activities performed during the experiment and objects directly involved or nearby.}
	\setlength{\tabcolsep}{12pt}
	\begin{tabular}{lll}
		\toprule
		activities		& objects involved			& objects nearby			\\ 
		\midrule
		open door		& door knob 				& door  					\\
		sitting			& chair or couch			& body, chair, couch, table	\\
		lying			& couch						& body, couch, cushion		\\
		get things		& plate, glass, cutlery, 	& shelf, locker, fridge, table \\ 
						& bread, marmalade, bottle  & 							\\
		making bread	& bread, knife, marmalade	& table, plate 				\\
		eating			& bread						& table, plate, body		\\
		drinking 		& bottle, glass 			& table, body				\\
		sleeping		& bed, cushion, blanket		& body 						\\
		walking			& 							& body						\\
		 \bottomrule
	\end{tabular}
	\label{tab:activities}
\end{table}

In order to evaluate if capacitive proximity sensors in wrist-bands can enhance the performance of activity recognition, we have conducted an evaluation with 7 test persons. All test persons received a basic plot with the activities they were supposed to perform. They were not given any instructions about the way they are supposed to perform the activities. After manual labeling, we used this test-set as ground truth and performed a 4-fold cross-validation on an support-vector machine (SVM) classifier on each user. The cross-validation was performed once with and once without including the data of the capacitive proximity sensor into the feature set. We chose an SVM classifier because of its high relevance in activity recognition and its fast performance.

@@

In the following, the performed activities will be analyzed in detail, stating out the influence of the capacitive proximity sensor on the classification result. In general, the usage of data provided by the new modality showed improvements in recognition rates reaching from 2.4 up to 10.7\% for a single activity.

\begin{figure}[h]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/tobias_1.pdf}
	\caption{When the participants entered the apartment, the wristband approached the door knob twice, at the time of opening and closing the door. This fact can be observed in the capacitive proximity data (upper plot) at the beginning and at the end of the activity, whereas the acceleration has no characteristic information (bottom plot).}
	\label{fig:opendoor}
\end{figure}

The opening door activity has very poor recognition rates without the data from the proximity sensor. The average F-measure could be increased from 35.5\% to 46.2\%. A plot of the activity is given in Figure \ref{fig:opendoor}. The capacitive proximity sensor shows two approaches to the door knob, one for opening the door (2s) and one for closing the door (9s). The acceleration sensor captures relevant data in the time in which the person moves into the room and the hand changes from the outer to the inner door knob (5 - 7s). The recorded data for this activity also shows strong correlations between all experiment participants. The confusion matrices show that the ``open door'' activity was often confused with the ``sitting'' activity, probably because of the amount of motion on the one hand and the proximity to nearby objects (door, couch or cushions) on the other hand. By using the capacitive sensor data, the recall for that class and confusion with the sitting activity could be improved.

@@

In the following, the participant were instructed to lie  down on the couch. Again, there were great variations in how this activity was performed by the participants. For example, some of them crossed their hands under their head, or placed them on their body. For this class, the average F-measure could only be increased by 2.4\%, from 81.2 to 83.6\%. Considering some participants, the activity was often confused with the ``making bread'' class. By using the proximity modality, the confusion between the two classes could be reduced.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/tobias_4.pdf}
	\caption{An example of the ``getting things'' activity, where the participants had to get food and dishes from shelves and lockers. The proximity sensor peaks in the beginning (9s and 19s) indicate immediate proximity to shelf, and to the locker (55-63s) in the kitchen (upper plot). The signal drop at the end results from the participant placing his hand on the table when she was finished.}
	\label{fig:get_things}
\end{figure}

After that, the participants were asked to walk over to the kitchen and to put food and dishes from a shelf and a locker on the table. This activity involved direct interactions with various objects as well as proximity to furniture in the room (see Table \ref{tab:activities}). The capacitive proximity sensor was able to capture the proximity to the shelf and to the table. The average F-measure for this activity is rather low, but improved by 6.8\% from 53.8\% to 60.6\%. 
The worst performing participants for this activity reached an F-measure of 46.8\% without and 53.5\% with the capacitive sensor, while the best performing one reached 62.0\% and 64.5\% respectively. The low performance results from confusions with other activities, with a higher tendency to the ``prepare bread'' activity across all participants. This is most likely due to various objects involved in both activities. The capacitive sensor modality has a much more positive impact reducing the confusion with other activities.

\begin{figure}[t]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a slice of bread. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows recurring hand motions.}
	\label{fig:prep_bread}
\end{figure}

Figure \ref{fig:prep_bread} shows an example instance of preparing a bread with marmalade. It is notable that the acceleration data does not seem to provide any characteristic patterns, while the proximity sensor indicates a table, plate, or other objects in immediate distance. This activity showed a high improvement in the average F-measure by 10\%, from 49.0\% to 59.8\%, where the data delivered by the capacitive proximity sensor is taken into account. The making bread class was often confused with the sitting class for some users, probably due to lots of motion during the sitting, as mentioned previously. For other users, making bread was confused with eating or drinking. Using the new input modality, confusion across users could be reduced.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An example of a participant eating a slice of bread with marmalade, taking 5 bites from it. After each bite, the hand is placed on the table, which can be recognized both in the acceleration as well as the proximity data plots.}
	\label{fig:eating}
\end{figure}

When considering the eating activity, the impact of the new capacitive proximity sensor on the classification performance is quite low, as the chosen features are able to distinguish it from other activities. Some of the participants ate their bread leaving their hand close to the mouth, while others moved their hand up and down putting their bread aside on the plate, which also results in the performance range from 71.2 to 88.1\% without and 77.5 to 90.6\% with the proximity data. An example of an eating activity is shown in Figure \ref{fig:eating}, where the participant took a few bites from the bread while putting it down every time. The average F-measure for that activity increased slightly by 2.7\% (from 79.5\% to 82.2\%).

@@

The activity ``drinking water'' is depicted in Figure \ref{fig:drinking}. The participant took a few drinks from the glass, while leaving the hands lying on the table in between. These motions can be easily detected in the acceleration as well as proximity data. The accelerometer data shows that there periodic up- and down-movements while the capacitive proximity sensor delivers data that is associated to the proximity of the table. The average classification performance lies at 48.4\% without and 54.4\% with the proximity data taken into account, resulting in a gain of 6\%.

@@

Regarding the walking activity one can identify periodic changes in the acceleration as well as in the measured capacitance, illustrated in Figure \ref{fig:walking}. While walking, the capacitance between the wristband and the leg increases when the wristband is located close to the body and decrease when the wristband moves away. There were problems distinguishing this activity from ``get things'' that could be improved by using the new input modality. The classification improvement for this activity accounts to 12.2\% boosting the average F-measure from 41.7\% without to 53.9\% with the new sensor. Due to the low performance results and the characteristic periodic signal shape, it would help to consider frequency domain features, as it is often applied in related work.

@@

The sleeping activity (an example shown in Figure \ref{fig:sleeping}) was classified with an average F-measure of 83.2\%, which increased to 86.9\% when using the proximity data. In this case, the capacitive proximity sensor is able to capture the surrounding cushions and blankets, as well as the body or head of the participant. The accelerometer data and the proximity data has larger periods with a constant signal, which is a cause for confusion with the ``lying on the couch'' activity.

\begin{figure}
	\setlength{\tabcolsep}{2pt}
	\centering
	\begin{scriptsize}
	\begin{tabular}{|ccccccccc|p{0.01cm}|ccccccccc|p{0.01cm}|l}
% 		\toprule
		\multicolumn{9}{c}{without proximity data} &\multicolumn{1}{c}{}& 
			\multicolumn{9}{c}{with proximity data} &\multicolumn{1}{c}{}&  \\ \cmidrule{1-9} \cmidrule{11-19}

		\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} &\multicolumn{1}{c}{}& 
			\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} 	&\multicolumn{1}{c}{}& $\leftarrow$ classified as\\
		\cline{1-9} \cline{11-19} \cline{21-21}
\textbf{8} & 0 & 0 & 0 & 0 & 0 & 0 & 2 & 1		&& \textbf{11} & 0 & 0 & 2 & 1 & 0 & 0 & 0 & 1 	&& a = open door	\\ 
0 & \textbf{140} & 0 & 5 & 2 & 0 & 0 & 0 & 0  	&& 0 & \textbf{140} & 0 & 5 & 2 & 0 & 0 & 0 & 0 && b = sitting		\\ 
0 & 0 & \textbf{294} & 1 & 9 & 10 & 1 & 1 & 0 	&& 0 & 0 & \textbf{309} & 1 & 0 & 4 & 0 & 0 & 1 && c = lying		\\ 
1 & 7 & 1 & \textbf{50} & 17 & 9 & 1 & 4 & 1	&& 2 & 6 & 0 & \textbf{67} & 14 & 9 & 1 & 2 & 1 && d = get things	\\ 
0 & 2 & 7 & 7 & \textbf{102} & 42 & 5 & 0 & 2	&& 1 & 1 & 0 & 13 & \textbf{120} & 24 & 7 & 1 & 1&& e = make bread	\\ 
0 & 0 & 3 & 5 & 21 & \textbf{288} & 4 & 0 & 0 	&& 0 & 0 & 2 & 1 & 19 & \textbf{300} & 3 & 0 & 0 && f = eating		\\ 
0 & 5 & 11 & 1 & 29 & 40 & \textbf{18} & 0 & 0	&& 0 & 4 & 4 & 0 & 26 & 57 & \textbf{16} & 0 & 0 && g = drinking	\\
2 & 0 & 0 & 2 & 0 & 0 & 0 & \textbf{11} & 0	 	&& 2 & 0 & 0 & 0 & 0 & 0 & 0 & \textbf{14} & 0 	 && h = walking		\\
0 & 2 & 2 & 0 & 2 & 0 & 0 & 0 & \textbf{271}	&& 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{276}  && i = sleeping	\\ 
		\cline{1-9} \cline{11-19} 
% 		\bottomrule
	\end{tabular}
	\end{scriptsize}

@@

A better classification performance can also be observed for the ``get things'' class that includes interactions with a multitude of objects in the environment. Regarding the activity ``make bread'', the capacitive proximity could reduce the number of confusions with the ``eating'' class significantly.

Due to the high similarity of eating and drinking (cf. Figure \ref{fig:eating} and \ref{fig:drinking}), the number of confusions between those two classes increases when considering the proximity data. However, for the ``drinking'' class the new modality limits the confusions to related activities such as ``eating'' and ``make bread'' only, while lowering the number of false recognitions for the other activities.

@@

Our paper presented a wrist-worn activity data logger prototype, which consists of an accelerometer in combination with a capacitive proximity sensor integrated into the wristband. Our experiments with seven participants and nine basic daily activities show that this additional input modality can significantly boost the activity recognition performance.
Regarding the classification performance, we obtained an improvement in the average F-measure of 6.3\%, from 67.2 to 73.5\%. Specifically, the activity classes ``walking'', ``preparing bread'' and ``open door'' could benefit a lot from proximity-related sensor data. For such classes, the classification performance could be boosted by 12.2, 9.0 and 10.8\% respectively.

With this proof of concept we show that the proximity information can provide an information gain regarding the evaluated activities. In future work we aim at evaluating other relevant feature types, such as frequency domain features, as well as feature sets to extract the most discriminative ones. Additionally, using other classifiers (such as HMMs) or more fitting classifier configurations might also improve activity recognition performance.

@@

\bibliographystyle{splncs03}
\bibliography{references}

>>>>

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{
 	Fraunhofer IGD, Darmstadt, Germany\\
 	\email{tobias.grosse-puppendahl@igd.fraunhofer.de}
 	\and
 	Embedded Sensing Systems, Technische Universit\"at Darmstadt, Germany\\
 	\email{\{berlin,borazio\}@ess.tu-darmstadt.de}
}

@@

\maketitle
\begin{abstract}
Activity recognition with a wearable accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the user's environment to boost activity recognition for more challenging activities. This work presents a sensor prototype consisting of an accelerometer and a capacitive proximity sensor that senses the user's activities based on the combined sensor values. We show that our proposed approach of combining both modalities significantly improves the recognition rate for detecting activities of daily living.
\keywords{activity recognition, capacitive proximity sensors, ambient assisted living, user context}
\end{abstract}

@@

Sensing a person's activity is an active research topic with a raising interest due to the advancement in mobile phone technology. These devices include multiple sensors and therefore enable the recognition of daily activities \cite{brezmes2009activity}. Current wearable activity recognition systems are able to unobtrusively capture and recognize a person's activities throughout the whole day. These systems often rely on inertial sensor data that are captured by wearable sensors embedded in a mobile device \cite{brezmes2009activity} or attached to the body \cite{Ravi2005}. Usually, single sensor modalities are used or duplicated to detect the activities. However, it is a great challenge to identify many activities just by using a single modality like the accelerometer.

Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the object's distance, its size and the material it is made of. In this work, we show that an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living that rely heavily upon object usage. Therefore, we obtained an open-hardware and open-source wrist-worn activity data logger \cite{hedgehog} and integrated a capacitive proximity sensor into its wristband.

There are several intuitive examples for which a combination of accelerometer-based activity recognition with a capacitive proximity sensor reveals its strength. For example, it may be conducted which material is placed underneath the wristband. The capacitive proximity sensor would return a different measurement result for a hand placed on a couch covered with fabric than a hand placed on a wooden table. Moreover, the approximate distance from the wristband to objects can be exploited to identify activities like grasping into a locker or a refrigerator to prepare food.

@@

A great variety of capacitive sensors and measurement techniques exists \cite{Smith1999}. The most common sensing principle, the loading mode, is based on running numerous charge and discharge cycles of the virtual capacitor that is created by the electrode and the environment. Depending on the charge and discharge times, one can infer the corresponding capacitance. This sensing principle is applied by Wimmer et al. in \cite{Wimmer2007} who presented a toolkit for capacitive proximity sensing. In previous works, capacitive proximity sensors were applied in various fields of human-computer-interaction. Wimmer et al. and Grosse-Puppendahl et al. presented gesture recognition systems \cite{Wimmer,Grosse-puppendahl2012} as well as smart furniture that can sense human activities \cite{Wimmer} and classify human postures \cite{Grosse-puppendahl2011}. Cheng et al. have investigated the possibility of using capacitive sensors for activity recognition by measuring shape changes of muscles and skin \cite{Cheng2010}. 
%To our knowledge, capacitive proximity sensors have not been embedded into a wearable device to enhance the performance of activity recognition.

@@

The sensor node itself is built around the low-power Microchip microcontroller (PIC18F46J50) featuring an accelerometer sensor (ADXL345) to capture human motion, light and ambient temperature sensors and a microSD flash card for locally storing the sensor data. The sensor is powered by a $200mAh$ lithium polymer battery, which allows for two weeks of continuous recording on a single battery charge. A USB port is used to configure the sensor (e.g. setting the sensitivity of the accelerometer), to access the stored sensor data, and to recharge the battery. A plastic case packages and protects the sensor to be worn at the wrist (Figure \ref{fig:sensornode}).

@@

The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode. The sensor draws a supply current of 1mA at 3.3V when active which qualifies it for wearable proximity sensing applications. In the following a virtual capacitor denotes the capacitance between the sensing electrode and the environment. The capacitance of the sensing electrode to environmental objects increases with closer distances.
% The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode that aims to eliminate the influence of the underlying skin. Without shielding no meaningful measurements could be performed, since the arm's skin is a very proximate grounded object to the sensing electrode. The sensor draws a supply current of 1mA at 3.3V when active and qualifies it for wearable proximity sensing applications. In the following a virtual capacitor denotes the capacitance between the sensing electrode and the environment. The capacitance of the sensing electrode to environmental objects increases with closer distances.

@@

The operations required for a measurement cycle are illustrated in Figure \ref{fig:pseudocode}. The proximity sensor board generates a clock signal with varying frequency depending on the charge and discharge cycles. The HedgeHog measures the resulting capacitance by counting the signal's edges over a gate time of approximately 9.5ms. During that counting phase, the microcontroller is sent to sleep in order to reduce power consumption. 
%In the following, the HedgeHog applies run-length encoding on the measured data to reduce overhead and periodically logs the data to the integrated microSD card.

@@

The recorded test set contains the following activities: opening door, sitting on a couch, lying on a couch, putting kitchen equipments from a shelf and out of a locker, making a marmalade sandwich, eating the sandwich, pouring and drinking water, walking and sleeping. The relations of these activities to environmental objects are given in Table \ref{tab:activities}. Some of those activities are very hard to recognize when the data is limited to a single modality like a 3D accelerometer. For example, sitting at the table and sitting on a couch are very similar activities. We aim to show that the data basis can be significantly improved by the additional input modality.

\begin{table}
	\centering
	\caption{Some details on the activities performed during the experiment and objects directly involved or nearby.}
	\setlength{\tabcolsep}{12pt}
	\begin{tabular}{lll}
		\toprule
		activities		& objects involved			& objects nearby			\\ 
		\midrule
		open door		& door knob 				& door  					\\
		sitting			& chair or couch			& body, chair, couch, table	\\
		lying			& couch						& body, couch, cushion		\\
		get things		& plate, glass, cutlery, 	& shelf, locker, fridge, table \\ 
						& bread, marmalade, bottle  & 							\\
		make sandwich	& bread, knife, marmalade	& table, plate 				\\
		eating			& marmalade sandwich		& table, plate, body		\\
		drinking 		& bottle, glass 			& table, body				\\
		sleeping		& bed, cushion, blanket		& body 						\\
		walking			& 							& body						\\
		 \bottomrule
	\end{tabular}
	\label{tab:activities}
\end{table}

In order to evaluate if capacitive proximity sensors in wrist-bands can enhance the performance of activity recognition, we have conducted an evaluation with 7 test persons. All test persons received a basic script with the activities they were supposed to perform. They were not given any instructions about the way they are supposed to perform the activities. After manual labelling, we used this test-set as ground truth and performed a 4-fold cross-validation on an support-vector machine (SVM) classifier on each user. The cross-validation was performed once with and once without including the data of the capacitive proximity sensor into the feature set. We chose an SVM classifier because of its high relevance in activity recognition and its fast performance.

@@

In the following, the performed activities will be analyzed in detail, stating the influence of the capacitive proximity sensor on the classification result. In general, the usage of data provided by the new modality showed improvements in recognition rates reaching from 2.4\% up to 10.7\% for single activities.

\begin{figure}[h]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/tobias_1.pdf}
	\caption{When the participants entered the apartment, the wrist approached the door knob twice, at the time of opening and closing the door. This fact can be observed in the capacitive proximity data (upper plot) at the beginning and at the end of the activity, whereas the acceleration has little characteristic information (bottom plot).}
	\label{fig:opendoor}
\end{figure}

The ''opening door'' activity has very poor recognition rates without the data from the proximity sensor. The average F-measure could be increased from 35.5\% to 46.2\%. A plot of the activity is given in Figure \ref{fig:opendoor}. The capacitive proximity sensor shows two approaches to the door knob, one for opening the door (2s) and one for closing the door (9s). The acceleration sensor captures relevant data in the time in which the person moves into the room and the hand changes from the outer to the inner door knob (5 - 7s). The recorded data for this activity also shows strong correlations between all experiment participants. The confusion matrices show that the ``open door'' activity was often confused with the ``sitting'' activity, probably because of the amount of motion on the one hand and the proximity to nearby objects (door, couch or cushions) on the other hand. By using the capacitive sensor data, the recall for that class and confusion with the sitting activity could be improved.

@@

In the following, the participant were instructed to lie  down on the couch. Again, there were great variations in how this activity was performed by the participants. For example, some of them crossed their hands under their head, or placed them on their body. For this class, the average F-measure could only be increased by 2.4\%, from 81.2 to 83.6\%. Considering some participants, the activity was often confused with the ``make sandwich'' class. By using the proximity modality, the confusion between the two classes could be reduced.

\begin{figure}[t]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/tobias_4.pdf}
	\caption{An example of the ``getting things'' activity, where the participants had to get food and dishes from shelves and lockers. The proximity sensor peaks in the beginning (9s and 19s) indicate immediate proximity to shelf, and to the locker (55-63s) in the kitchen (upper plot). The signal drop at the end results from the participant placing his hand on the table when she was finished.}
	\label{fig:get_things}
\end{figure}

After that, the participants were asked to walk over to the kitchen and to put food and dishes from a shelf and a locker on the table. This activity involved direct interactions with various objects as well as proximity to furniture in the room (see Table \ref{tab:activities}). The capacitive proximity sensor was able to capture the proximity to the shelf and to the table. The average F-measure for this activity is rather low, but improved by 6.8\% from 53.8\% to 60.6\%. 
The worst performing participants for this activity reached an F-measure of 46.8\% without and 53.5\% with the capacitive sensor, while the best performing one reached 62.0\% and 64.5\% respectively. The low performance results from confusions with other activities, with a higher tendency to the ``make sandwich'' activity across all participants. This is most likely due to various objects involved in both activities, and the fact that 1-second features are obtained. The capacitive sensor modality has a much more positive impact reducing the confusion with other activities.

\begin{figure}[t]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``make sandwich'' activity, where the participants had to put marmalade on a slice of bread. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows recurring hand motions.}
	\label{fig:prep_bread}
\end{figure}

Figure \ref{fig:prep_bread} shows an example instance of preparing a bread with marmalade. It is notable that the acceleration data does not seem to provide any characteristic patterns, while the proximity sensor indicates a table, plate, or other objects in immediate distance. This activity showed a high improvement in the average F-measure by 10\%, from 49.0\% to 59.8\%, where the data delivered by the capacitive proximity sensor is taken into account. The ''make sandwich'' class was often confused with the ''sitting'' class for some users, probably due to lots of motion during the sitting, as mentioned previously. For other users, ''make sandwich'' was confused with ''eating'' or ''drinking''. Using the new input modality, confusion across users could be reduced.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An example of a participant eating a marmalade sandwich, taking 5 bites from it. After each bite, the hand is placed on the table, which can be recognized both in the acceleration as well as the proximity data plots.}
	\label{fig:eating}
\end{figure}

When considering the eating activity, the impact of the new capacitive proximity sensor on the classification performance is quite low, as the chosen features are able to distinguish it from other activities. Some of the participants ate their sandwich leaving their hand close to the mouth, while others moved their hand up and down putting their sandwich aside on the plate, which also results in the performance range from 71.2 to 88.1\% without and 77.5 to 90.6\% with the proximity data. An example of an eating activity is shown in Figure \ref{fig:eating}, where the participant took a few bites from the sandwich while putting it down every time. The average F-measure for that activity increased slightly by 2.7\% (from 79.5\% to 82.2\%).

@@

The activity ``drinking water'' is depicted in Figure \ref{fig:drinking}. The participant took a few sips from the glass, with leaving the hands positioned. These motions can be easily detected in the acceleration as well as proximity data. The accelerometer data shows that there periodic up- and down-movements while the capacitive proximity sensor delivers data that is associated to the proximity of the table. The F-measure lies at 48.4\% without and 54.4\% with the proximity data taken into account, resulting in a gain of 6\%.

@@

Regarding the walking activity one can identify periodic changes in the acceleration as well as in the measured capacitance, illustrated in Figure \ref{fig:walking}. While walking, the capacitance between the wristband and the leg increases when the wristband is located close to the body and decreases when the wristband moves away. There were problems distinguishing this activity from ``get things'' that could be improved by using the new input modality. The classification improvement for this activity accounts to 12.2\% boosting the average F-measure from 41.7\% without to 53.9\% with the new sensor. Due to the low performance results and the characteristic periodic signal shape, it would help to consider frequency domain features, as it is often applied in related work.

@@

The sleeping activity (an example shown in Figure \ref{fig:sleeping}) was classified with an average F-measure of 83.2\%, which increased to 86.9\% when using the proximity data. In this case, the capacitive proximity sensor is able to capture the surrounding cushions and blankets, as well as the body or head of the participant. The accelerometer data and the proximity data have larger periods of low variance, which is a cause for confusion with the ``lying on the couch'' activity.

\begin{figure}[t]
	\setlength{\tabcolsep}{1.7pt}
	\centering
	\begin{scriptsize}
	\begin{tabular}{|ccccccccc|p{0.01cm}|ccccccccc|p{0.01cm}|l}
% 		\toprule
		\multicolumn{9}{c}{without proximity data} &\multicolumn{1}{c}{}& 
			\multicolumn{9}{c}{with proximity data} &\multicolumn{1}{c}{}&  \\ \cmidrule{1-9} \cmidrule{11-19}

		\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} &\multicolumn{1}{c}{}& 
			\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} 	&\multicolumn{1}{c}{}& $\leftarrow$ classified as\\
		\cline{1-9} \cline{11-19} \cline{21-21}
\textbf{8} & 0 & 0 & 0 & 0 & 0 & 0 & 2 & 1		&& \textbf{11} & 0 & 0 & 2 & 1 & 0 & 0 & 0 & 1 	&& a = open door	\\ 
0 & \textbf{140} & 0 & 5 & 2 & 0 & 0 & 0 & 0  	&& 0 & \textbf{140} & 0 & 5 & 2 & 0 & 0 & 0 & 0 && b = sitting		\\ 
0 & 0 & \textbf{294} & 1 & 9 & 10 & 1 & 1 & 0 	&& 0 & 0 & \textbf{309} & 1 & 0 & 4 & 0 & 0 & 1 && c = lying		\\ 
1 & 7 & 1 & \textbf{50} & 17 & 9 & 1 & 4 & 1	&& 2 & 6 & 0 & \textbf{67} & 14 & 9 & 1 & 2 & 1 && d = get things	\\ 
0 & 2 & 7 & 7 & \textbf{102} & 42 & 5 & 0 & 2	&& 1 & 1 & 0 & 13 & \textbf{120} & 24 & 7 & 1 & 1&& e = make sandwich	\\ 
0 & 0 & 3 & 5 & 21 & \textbf{288} & 4 & 0 & 0 	&& 0 & 0 & 2 & 1 & 19 & \textbf{300} & 3 & 0 & 0 && f = eating		\\ 
0 & 5 & 11 & 1 & 29 & 40 & \textbf{18} & 0 & 0	&& 0 & 4 & 4 & 0 & 26 & 57 & \textbf{16} & 0 & 0 && g = drinking	\\
2 & 0 & 0 & 2 & 0 & 0 & 0 & \textbf{11} & 0	 	&& 2 & 0 & 0 & 0 & 0 & 0 & 0 & \textbf{14} & 0 	 && h = walking		\\
0 & 2 & 2 & 0 & 2 & 0 & 0 & 0 & \textbf{271}	&& 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{276}  && i = sleeping	\\ 
		\cline{1-9} \cline{11-19} 
% 		\bottomrule
	\end{tabular}
	\end{scriptsize}

@@

A better classification performance can also be observed for the ``get things'' class that includes interactions with a multitude of objects in the environment. Regarding the activity ``make sandwich'', the capacitive proximity could reduce the number of confusions with the ``eating'' class significantly.

Due to the high similarity of eating and drinking (cf. Figure \ref{fig:eating} and \ref{fig:drinking}), the number of confusions between those two classes increases when considering the proximity data. However, for the ``drinking'' class the new modality limits the confusions to related activities such as ``eating'' and ``make sandwich'' only, while lowering the number of false recognitions for the other activities.

@@

Our paper presented a wrist-worn activity data logger prototype, which consists of an accelerometer in combination with a capacitive proximity sensor integrated into the wristband. Our experiments with seven participants and nine daily activities show that this additional input modality can significantly boost the activity recognition performance.
Regarding the classification performance, we obtained an improvement in the average F-measure of 6.3\%, from 67.2 to 73.5\%. Specifically, the activity classes ``walking'', ``make sandwich'' and ``open door'' could benefit a lot from proximity-related sensor data. For such classes, the classification performance could be boosted by 12.2\%, 9.0\% and 10.8\% respectively.

With this proof of concept we show that the proximity information can provide an information gain regarding the evaluated activities. In future work we aim at evaluating other relevant feature types, such as frequency domain features, as well as feature sets to extract the most discriminative ones. Additionally, using other classifiers (such as HMMs) might also improve activity recognition performance.

@@

\bibliographystyle{splncs03}
%\bibliography{references}
